\textsc{LiveRec}: Framing Debug Protocols to Hotwire Probes

### 2023-09-15 paper/paper.tex
Continued reviewing. Added comments until Section 6.
Continued reviewing. Added comments until Section 7.
Parked ideas about the relationship between stack recordings, cause-and-effect chains and different live programming features.

A live environment must be able to react to two different events: a change in the code or a change in the test/input data. 
%vRozen: 'live environment' -- 'live programming environment'
If the code changes, we need to re-execute the code, and in the case of a compiled language, we need to compile the new code first. 
%vRozen: who is 'we' ? Consider replacing 'we' by 'a programmer' Whenever a programmer changes the code, ...
If the data changes, we need to be able to re-execute the programme with the new data. 
%vRozen: Likewise, the program needs to re-execute immediately when the programmer adjusts the inputs.
In addition, it is necessary to keep compilation and execution times low enough to maintain an interactive experience with the user.
%vRozen: 'the user' -- suggest 'the programmer' or 'programmers'

@@

\begin{figure}[h]
  \centering
  \begin{minipage}{0.25\textwidth}
    \centering
    \begin{lstlisting}[language=C]
//foo(3)
int foo(int n) {
  int i = 0;
  while (i < n) {
    i++;
  }
  return i;
}
    \end{lstlisting}
  \end{minipage}
  \hfill
  \begin{minipage}{0.7\textwidth}
    \centering
    \includegraphics[width=0.8\linewidth]{img/stackrecording_chain.png}
  \end{minipage}
  \caption{Stack Recording Example}
  \label{fig:stack-recording}
\end{figure}
%vRozen. The layout seems a bit off.

@@

\begin{figure}
  \centering
  \large  
  \hspace{8pt} \color{darkgray} \textsf{\textbf{time}}  \color{black} $\rightarrow$ \raggedright
  \vspace{2pt}

@@

In order to generate the data needed to probe a variable, we need to be able to retrieve the state of the variable during execution. 
This information is contained in the stackframe at the time of execution. 
However, we need to add spatial and temporal information to this: we need to associate each stackframe state with the location in the code at which it was retrieved, and we also need to know the order in which these stackframes were retrieved.

In this paper, we introduce a structure for representing this data: a \textit{stack recording}. 
A stack record represents the different states of the stack frame during the execution of a method. The stack record is presented as a chain of recorded stack frames, to which information has been added about the source code's location and height in the stack.
This representation allows us to maintain a link between the spatial location (the reference to the source code) and the temporal location (the order of these stackframes) of the execution.
This representation has several advantages: it is easy to construct from debugger information, and it applies to most programming languages.
Figure \ref{fig:stack-recording} shows an example of a stack recording. On the left is a C function and on the right is the stack recording of the execution of \code{foo(3)}. Each rectangle represents the state of the stackframe during execution, with the source code location (here simplified to the line number).

\begin{figure}[h]
  \centering
  \includegraphics[width=0.6\linewidth]{img/stackrecording_impl.png}
  \caption{Record Stackrecording (placeholder)}
  \label{fig:stack-recording-impl}
\end{figure}

Figure \ref{fig:stack-recording-impl} shows how a stack recording is recorded.
To obtain the different states of the stackframe during the execution of a method, we place a breakpoint at the start of the function. 
Once this has been reached, the method is executed step by step, using the debugger's step instruction. 
Between each step, the state of the stackframe is recorded along with the source code's location information.

@@

To enable stack recording, we developed a debugger leveraging the Java Debug Interface (JDI) library. 
A simplified pseudo-code representation of this debugger is displayed in Figure \ref{fig:debugger-class}. 
The class contains a method for starting the JVM (at line 2), a process to load or reload a class within the VM (at line 5), and a procedure for generating a method stack record (at line 10). Although in practice, the implementation involves multiple interacting threads, these details have been omitted from this overview in favor of clarity and simplicity.

@@

Class loading is achieved using a dynamic ClassLoader, which gives us the ability to append class paths and load a class during execution. 
Modifications to the code cause the class to be reloaded into the KeepAliveAgent. 
However, there is a caveat to this procedure - it only works if the class signature remains consistent and only the method content is changed. 
Any changes to the structure will require a restart of the KeepAliveAgent. 
It is to be noted that specific JVM implementations can bypass this limitation \cite{}.

To execute a function (line 10 in Debugger class, Figure \ref{fig:debugger-class}), the debugger creates the values associated with the arguments locally. 
It then transmits these values to the JVM containing the KAA using the \code{mirrorOf} and \code{newInstance} methods of JDI's Reflection API (line 11). 
The debugger then sets a breakpoint at the target method's input (line 12) and launches the target method using \code{invokeMethod}. 
The debugger then registers the stack using the algorithm described in the previous paragraph.
Until the method is completed, the debugger registers the stack frame and takes a step forward in debugging.

\section{Generalizing Live Probes with Debugger Adapter Protocol}
\label{sec:generalizing-live-probes}
The Debug Adapter Protocol (DAP), developed by Microsoft, is a standard method of communicating with a programming language debugger. It is compatible with various editors, including VS Code, and provides a unified interface for all programming languages.

By exploiting this protocol, we have created a new language parametric backend, which we have implemented for the C, Python, Java and Javascript programming languages. These languages are chosen to cover both compiled and interpreted languages. 
This backend offers an interface common to all three languages, which includes methods for starting and initialising the debugging server, loading and reloading code in the debugger, and executing a method while performing stack recording. 
Table \ref{tab:dap-req} contains a list of the requests from the DAP that we have used to implement the interface.

@@

These methods allow us to carry out stack recording independently of the chosen language and facilitate the future implementation of live programming interfaces.

The implementation for each language includes a keep-alive agent and code to communicate with the debugger and keep-alive agent to provide the interface functionality.
These methods depend on both the implementation of the debugging server and the keep-alive agent. The differences between the implementations are shown in the Table \ref{tab:implementation-differences}.

@@

\subsection{Implementation specifications}

\begin{table}[!ht]
  %\centering
  \noindent\setlength\tabcolsep{4pt}%
  \begin{tabular}{lrlrl}
      \toprule
      Implementation & \multicolumn{2}{l}{\#SLOC Probe Server} & \multicolumn{2}{l}{\#SLOC Keep-Alive Agent} \\
      \midrule
      Java JDI & 521 & Java & 75 & Java \\ 
      DAP Base & 270  & Python & --- & \\ 
      DAP Python & 142 & Python & 26 & Python \\ 
      DAP Java & 360 & Python & 114 & Java \\ 
      DAP Javascript & 256 & Python & 24 & Javascript \\ 
      DAP C & 181 & Python & 14 & C \\
      \bottomrule
  \end{tabular}
  \caption{Size of the language specific components in SLOC and their implementation languages.}
  \label{tab:sloc}
\end{table}

The table \ref{tab:sloc} shows the number of lines of code for the server part and the KAA for each implementation, with the language in brackets. The DAP Base line corresponds to the common code used by the other DAP-based implementations.
The tests were carried out with OpenJDK 20.0.2 for Java, Python 3.11.2 for Python and Node 18.16.1 for Javascript. The C code was compiled using GCC 13.1.1 and GDB 13.2 was used for stack recording. The machine used to carry out the evaluations has an AMD Ryzen 5 2500U CPU and 8Gb of RAM.

@@

We have developed a dynamic programming environment that integrates live probes via Java, C, Python and Javascript debugging servers.

Each time changes are made in the code editor, the code is sent to the server, which then checks for parseability and syntax errors. 
In addition, if the code contains a comment beginning with "@", followed by a function call, the server attempts to create a stack recording of that function, with the parameters provided.

The user interface consists of a web page that interacts with a local server. The demo features two display methods.

@@

The first mode is inspired by Bret Victor's presentation. The screenshot in Figure \ref{fig:demo} shows a session with a C binary search function. 
On the left side there is a code editor (based on CodeMirror 5), while on the right a panel displays live probes. The display on the right is generated from source code and stackrecording with a tree sitter to display probes in different locations:

\begin{itemize}
  \item Display of input arguments and return values for each function call.
  \item Displays the value of each variable definition or assignment. In cases where the assignment is in a loop or is called several times, all the different values are displayed.
  \item Display comparison values for each loop (for and while).
\end{itemize}

The use of the tree-sitter is not required to display probes. In fact, it could simply display the variables when they are modified. However, using the tree sitter also allows us to be able to display the state of variables where they are used, such as in comparisons in loops.

@@

The second mode, featured in Figure \ref{fig:demo-stack}, displays the stack recording in the form of a history. The screenshots show the different stack states recorded. The right-hand panel shows the stackframe state and a slider for exploring the different states. The location in the source code corresponding to the state where the stackframe has been registered is highlighted in yellow in the Codemirror editor on the left.

@@

In the context of a live programming environment, it is essential to have short response times after user interactions. 
We therefore measured the performance of our approach in response to various questions:

\begin{itemize}
  \item How long does it take to compile and to load the code into the debugger, depending on the size of the programme?
  \item How long does it take to run the program, depending on the number of stack frames needed to make the stack recording?
  \item What is the average performance during a live programming session?
\end{itemize}

\subsubsection{How long does it take to compile and to load the code into the debugger, depending on the size of the programme?}

@@

In our evaluation, we assessed the time required to compile and load code into the debugger memory for Python, C, Java and Javascript. 
To accomplish this, we compiled and loaded programs ranging from 5 to 100 lines of code. 
The summarized results can be found in Figure \ref{fig:compileload}.

@@

\subsubsection{How long does it need to run the program, depending on the number of steps to make the stack recording ?}

@@

Next, we evaluated the execution time as a function of the number of steps in the debugger. 
Figure \ref{fig:execute-time} shows the total time taken as a function of the number of steps during stack registration.
Note that the execution time shows a linear relationship with the number of steps in the method. 
In fact, for each implementation we obtain a constant time for each stack frame recorded, corresponding to the time taken to obtain the state of the stack frame and the time taken for the debugger to perform a step. 
The difference in time between the different DAP languages is explained by the implementation of the debug server. 
With C, the debug server is a simple wrapper around gdb, which has very good performance.
Figure \ref{fig:execute-java-jdi} shows Java execution times with DAP and JDI. 
The y-axis is in logarithmic scale. 
There is a significant difference in performance between the two implementations, despite the fact that the DAP server implementation in Java also uses JDWP.
By profiling the execution of the java debug server, we found two causes for this performance gap:
\begin{itemize}
  \item The first is due to the "evaluate" request in the DAP protocol. To run the function and define the arguments in the DAP implementation, we use "evaluate" requests. These requests require an additional compilation step in the debug server compared with the version using JDI. In the JDI version, the arguments are actually copied into the debugger directly from the server coded in Java. To improve the performance of our DAP implementation, if the arguments are not changed between two executions, they are kept in the KAA.
  \item The second cause is the native performance of the debugger server. Our method automates debugger procedures, and the observed performance of these procedures is sufficient for human use. For example, obtaining the stacktrace takes an average of 40 ms. When used by a person, this delay is less than 0.1 seconds and is considered instantaneous. However, when we perform stack recording, this request will be executed a significant number of times (depending on the size of the code).
\end{itemize}

\subsubsection{What is the average performance during a live programming session?}

@@

To test the performance of our approach under real conditions, we recreated the live programming scenario presented in the video by bret victor. 
The scenario describes the programming of a function performing a binary search on an array of characters. It consists of 19 steps, including 10 steps where the source code is modified and 9 steps where the parameters are changed.

@@

>>>>

A live environment must be able to react to two different events: a change in the code or a change in the test/input data. 
%vRozen: 'live environment' -- 'live programming environment'
If the code changes, we need to re-execute the code, and in the case of a compiled language, we need to compile the new code first. 
%vRozen: who is 'we' ? Consider replacing 'we' by 'a programmer' Whenever a programmer changes the code, ...
%vRozen: alternatively, 'we' could be software language engineers who are implementing probes for language X.
If the data changes, we need to be able to re-execute the programme with the new data. 
%vRozen: Likewise, the program needs to re-execute immediately when the programmer adjusts the inputs.
In addition, it is necessary to keep compilation and execution times low enough to maintain an interactive experience with the user.
%vRozen: 'the user' -- suggest 'the programmer' or 'programmers'

@@

\begin{figure}[h]
  \centering
  \begin{minipage}{0.25\textwidth}
    \centering
    \begin{lstlisting}[language=C]
//foo(3)
int foo(int n) {
  int i = 0;
  while (i < n) {
    i++;
  }
  return i;
}
    \end{lstlisting}
  \end{minipage}
  \hfill
  \begin{minipage}{0.7\textwidth}
    \centering
    \includegraphics[width=0.8\linewidth]{img/stackrecording_chain.png}
  \end{minipage}
  \caption{Stack Recording Example}
  \label{fig:stack-recording}
\end{figure}
%vRozen. The layout seems a bit off.
%vRozen. I propose merging Figure 3 and Table 1 into a single figure, the tabularx version below.

@@

%vRozen: Perhaps removing the time/space arrows would be better. it seems a bit cluttered
%vRozen: Perhaps we can still add boxes around the stack recordings to make it more similar to figure 4. It gives the feeling of having data structures, not just values.
\begin{figure}
  \centering
  \large  
  \hspace{8pt} \color{darkgray} \textsf{\textbf{time}}  \color{black} $\rightarrow$ \raggedright
  \vspace{2pt}

@@

In order to generate the data needed to probe a variable, we need to be able to retrieve the state of the variable during execution. 
%vRozen: Repeats 'need' -- consider rephrasing
This information is contained in the stackframe at the time of execution. 
%vRozen: passive -- 'contains' is more active
However, we need to add spatial and temporal information to this: we need to associate each stackframe state with the location in the code at which it was retrieved, and we also need to know the order in which these stackframes were retrieved.
%vRozen: Repeats 'we need' -- Long sentence. Consider breaking the sentence into parts.
%vRozen: Probing a variable requires both spacial and temporal information. ...

In this paper, we introduce a structure for representing this data: a \textit{stack recording}. 
A stack record represents the different states of the stack frame during the execution of a method. The stack record is presented as a chain of recorded stack frames, to which information has been added about the source code's location and height in the stack.
%vRozen: Repeats 'stack record' and repeats the part of 'we need' in different words.
This representation allows us to maintain a link between the spatial location (the reference to the source code) and the temporal location (the order of these stackframes) of the execution.
This representation has several advantages: it is easy to construct from debugger information, and it applies to most programming languages.
Figure \ref{fig:stack-recording} shows an example of a stack recording. On the left is a C function and on the right is the stack recording of the execution of \code{foo(3)}. Each rectangle represents the state of the stackframe during execution, with the source code location (here simplified to the line number).

\begin{figure}[h]
  \centering
  \includegraphics[width=0.6\linewidth]{img/stackrecording_impl.png}
  \caption{Record Stackrecording (placeholder)}
  \label{fig:stack-recording-impl}
\end{figure}
%vRozen: This is a UML sequence diagram.
%vRozen: The loop may use an activation frame.
%vRozen: The objects (or component instances) on the top of life lines appear in rectangular boxes.

Figure \ref{fig:stack-recording-impl} shows how a stack recording is recorded.
%vRozen: Repetitive 'stack recording' and 'recorded', consider rephrasing.
To obtain the different states of the stackframe during the execution of a method, we place a breakpoint at the start of the function.
%vRozen: method or function? -- pick one, at the beginning. Perhaps state that in this paper we can use these terms interchangeably.
Once this has been reached, the method is executed step by step, using the debugger's step instruction.
%vRozen: Repeats: step step step. Passive sentence. One the execution reaches this break point, it continues in a stepwise manner.
Between each step, the state of the stackframe is recorded along with the source code's location information.

@@

To enable stack recording, we developed a debugger leveraging the Java Debug Interface (JDI) library. 
%vRozen: Past tense. Please use present tense whenever possible for simplicity. Use past perfect if necessary, stressing a solution has addressed a challenge.
A simplified pseudo-code representation of this debugger is displayed in Figure \ref{fig:debugger-class}. 
%vRozen. Passive. Suggest a more active sentence. Figure .. displays a simplified pseudo-code representation of this debugger.
The class contains a method for starting the JVM (at line 2), a process to load or reload a class within the VM (at line 5), and a procedure for generating a method stack record (at line 10). 
Although in practice, the implementation involves multiple interacting threads, these details have been omitted from this overview in favor of clarity and simplicity.
%vRozen: High mental load for the reader. Suggest adding some signals to the text. For conciseness we omit <<thing>>. In practice, yada yada <<additional complexity>>.

@@

Class loading is achieved using a dynamic ClassLoader, which gives us the ability to append class paths and load a class during execution. 
%vRozen: passive sentence. A ClassLoader loads classes dynamically during the execution. By appending the class path ...

Modifications to the code cause the class to be reloaded into the KeepAliveAgent.
%vRozen: KAA, Keep Alive Agent or KeepAliveAgent?  
However, there is a caveat to this procedure - it only works if the class signature remains consistent and only the method content is changed. 
Any changes to the structure will require a restart of the KeepAliveAgent. 
It is to be noted that specific JVM implementations can bypass this limitation \cite{}.
%vRozen: heavy it. missing citation

To execute a function (line 10 in Debugger class, Figure \ref{fig:debugger-class}), the debugger creates the values associated with the arguments locally.
%vRozen: Whenever a probing session executes a function (line 10), the debugger creates ...

It then transmits these values to the JVM containing the KAA using the \code{mirrorOf} and \code{newInstance} methods of JDI's Reflection API (line 11).
%vRozen: KAA, Keep Alive Agent or KeepAliveAgent? 
The debugger then sets a breakpoint at the target method's input (line 12) and launches the target method using \code{invokeMethod}. 
The debugger then registers the stack using the algorithm described in the previous paragraph.
Until the method is completed, the debugger registers the stack frame and takes a step forward in debugging.
%vRozen: The above paragraph starts with a 'heavy it' followed by 'the debugger' several times. Perhaps use a "First, the debugger ... Next, it ... Finally, ..." temporal structure.

\section{Generalizing Live Probes with Debugger Adapter Protocol}
\label{sec:generalizing-live-probes}
The Debug Adapter Protocol (DAP), developed by Microsoft, is a standard method of communicating with a programming language debugger. It is compatible with various editors, including VS Code, and provides a unified interface for all programming languages. 
%vRozen: heavy it. Suggest rephrasing: DAP is ...

By exploiting this protocol, we have created a new language parametric backend, which we have implemented for the C, Python, Java and Javascript programming languages.
%vRozen: Only here it becomes clear 'us' is we the authors -- creators of generic language technology.
%vRozen: Long sentence.
These languages are chosen to cover both compiled and interpreted languages. 
%vRozen: Passive. We choose, ...
This backend offers an interface common to all three languages, which includes methods for starting and initialising the debugging server, loading and reloading code in the debugger, and executing a method while performing stack recording.
%vRozen: Long sentence, consider breaking into parts. 
Table \ref{tab:dap-req} contains a list of the requests from the DAP that we have used to implement the interface.

@@

These methods allow us to carry out stack recording independently of the chosen language and facilitate the future implementation of live programming interfaces. 
%vRozen: Short paragraph. Consider merging.

The implementation for each language includes a keep-alive agent and code to communicate with the debugger and keep-alive agent to provide the interface functionality.
%vRozen: Keep Alive Agent, KAA, KeepAliveAgent or keep-alive agent?
These methods depend on both the implementation of the debugging server and the keep-alive agent. 
%vRozen: These methods --> Specific details 
The differences between the implementations are shown in the Table \ref{tab:implementation-differences}.
%vRozen: Passive. Figure .. highlights the differences between the implementations.
%vRozen: Some connective sentence. We explain ...

@@

\subsection{Implementation specifications}
%vRozen: todo fixme -- chose different heading

\begin{table}[!ht]
  %\centering
  \noindent\setlength\tabcolsep{4pt}%
  \begin{tabularx}{\textwidth}{X r X r X}
      \toprule
      Implementation & \multicolumn{2}{l}{\#SLOC Probe Server} & \multicolumn{2}{l}{\#SLOC Keep-Alive Agent} \\
      \midrule
      Java JDI & 521 & Java & 75 & Java \\ 
      DAP Base & 270  & Python & --- & \\ 
      DAP Python & 142 & Python & 26 & Python \\ 
      DAP Java & 360 & Python & 114 & Java \\ 
      DAP Javascript & 256 & Python & 24 & Javascript \\ 
      DAP C & 181 & Python & 14 & C \\
      \bottomrule
  \end{tabularx}
  \caption{Size of the language specific components in SLOC and their implementation languages.}
  \label{tab:sloc}
\end{table}

The table \ref{tab:sloc} shows the number of lines of code for the server part and the KAA for each implementation, with the language in brackets. The DAP Base line corresponds to the common code used by the other DAP-based implementations.
The tests were carried out with OpenJDK 20.0.2 for Java, Python 3.11.2 for Python and Node 18.16.1 for Javascript. The C code was compiled using GCC 13.1.1 and GDB 13.2 was used for stack recording. The machine used to carry out the evaluations has an AMD Ryzen 5 2500U CPU and 8Gb of RAM.
%vRozen: Avoid complicated past perfect tense 'was compiled', e.g., 'has been compiled' or 'we have compiled'.

@@

We have developed a dynamic programming environment that integrates live probes via Java, C, Python and Javascript debugging servers. 
%vRozen: Unnecessary switch between perfect tense and simple present tense.

Each time changes are made in the code editor, the code is sent to the server, which then checks for parseability and syntax errors.
%vRozen: 'parseability' ? The server parses the code and checks for syntax errors.
In addition, if the code contains a comment beginning with "@", followed by a function call, the server attempts to create a stack recording of that function, with the parameters provided. 
%vRozen: Long sentence. Consider breaking into parts.
%vRozen: Repeats 'the server'. Perhaps use 'it'.

The user interface consists of a web page that interacts with a local server. The demo features two display methods. 
%vRozen: Short paragraph. Consider merging.

@@

The first mode is inspired by Bret Victor's presentation. The screenshot in Figure \ref{fig:demo} shows a session with a C binary search function. 
%vRozen: Figure X depicts screenshot of a session that executes a binary search function written in C.
On the left side there is a code editor (based on CodeMirror 5), while on the right a panel displays live probes. 
%vRozen: Consider rephrasing the part after the comma, specifically 'while' and 'live probes'.
The display on the right is generated from source code and stackrecording with a tree sitter to display probes in different locations: 
%vRozen: Long sentence. Consider breakin into parts. 'stack recording'

\begin{itemize}
  \item Display of input arguments and return values for each function call.
  \item Displays the value of each variable definition or assignment. In cases where the assignment is in a loop or is called several times, all the different values are displayed.
  \item Display comparison values for each loop (for and while).
\end{itemize}
%vRozen: repeats display(s). Consider saying it once, before the points.

The use of the tree-sitter is not required to display probes. In fact, it could simply display the variables when they are modified. However, using the tree sitter also allows us to be able to display the state of variables where they are used, such as in comparisons in loops.
%vRozen. The significance of tree-sitter (or tree sitter?) has not yet been established. What is it and why is it needed?

@@

The second mode, featured in Figure \ref{fig:demo-stack}, displays the stack recording in the form of a history. The screenshots show the different stack states recorded. The right-hand panel shows the stackframe state and a slider for exploring the different states. The location in the source code corresponding to the state where the stackframe has been registered is highlighted in yellow in the Codemirror editor on the left.
%vRozen: stack frame or stackframe.

@@

In the context of a live programming environment, it is essential to have short response times after user interactions. 
%vRozen: Short response times are essential for live programming environments.
We therefore measured the performance of our approach in response to various questions:

\begin{itemize}
  \item How long does it take to compile and to load the code into the debugger, depending on the size of the programme?
  \item How long does it take to run the program, depending on the number of stack frames needed to make the stack recording?
  \item What is the average performance during a live programming session?
\end{itemize}
%vRozen: it helps to be precise on what is recorded, e.g., time in ms. How is 'performance' defined -- the average duration of each entire case?
%vRozen: My advise is to make the 'protocol' or 'steps' more precise, such that it because straightforward to reproduce and compare the results. 
%vRozen: Which programs are assessed? Is this all about the binary search function or is it about more programs?

\subsubsection{How long does it take to compile and to load the code into the debugger, depending on the size of the programme?}
%vRozen: todo fixme Replace heading by a descriptive title.

@@

In our evaluation, we assessed the time required to compile and load code into the debugger memory for Python, C, Java and Javascript. 
To accomplish this, we compiled and loaded programs ranging from 5 to 100 lines of code. 
The summarized results can be found in Figure \ref{fig:compileload}.
%vRozen: Paragrap is in past tense.

@@

\subsubsection{How long does it need to run the program, depending on the number of steps to make the stack recording ?}
%vRozen: todo fixme Replace heading by a descriptive title.

@@

Next, we evaluated the execution time as a function of the number of steps in the debugger. 
%vRozen: Past tense.
Figure \ref{fig:execute-time} shows the total time taken as a function of the number of steps during stack registration.
Note that the execution time shows a linear relationship with the number of steps in the method.
In fact, for each implementation we obtain a constant time for each stack frame recorded, corresponding to the time taken to obtain the state of the stack frame and the time taken for the debugger to perform a step. 
%vRozen: Long sentence.
The difference in time between the different DAP languages is explained by the implementation of the debug server. 
With C, the debug server is a simple wrapper around gdb, which has very good performance.
Figure \ref{fig:execute-java-jdi} shows Java execution times with DAP and JDI. 
The y-axis is in logarithmic scale. 
There is a significant difference in performance between the two implementations, despite the fact that the DAP server implementation in Java also uses JDWP.
By profiling the execution of the java debug server, we found two causes for this performance gap:
\begin{itemize}
  \item The first is due to the "evaluate" request in the DAP protocol. To run the function and define the arguments in the DAP implementation, we use "evaluate" requests. These requests require an additional compilation step in the debug server compared with the version using JDI. In the JDI version, the arguments are actually copied into the debugger directly from the server coded in Java. To improve the performance of our DAP implementation, if the arguments are not changed between two executions, they are kept in the KAA.
  \item The second cause is the native performance of the debugger server. Our method automates debugger procedures, and the observed performance of these procedures is sufficient for human use. For example, obtaining the stacktrace takes an average of 40 ms. When used by a person, this delay is less than 0.1 seconds and is considered instantaneous. However, when we perform stack recording, this request will be executed a significant number of times (depending on the size of the code).
\end{itemize}
%vRozen: Consider using paragraphs instead of bullet points.

\subsubsection{What is the average performance during a live programming session?}
%vRozen: todo fixme Replace heading by a descriptive title.

@@

To test the performance of our approach under real conditions, we recreated the live programming scenario presented in the video by bret victor. 
%vRozen: I would expect to see this earlier, where i commented on the 'protocol'. The structure could have a more clear division between describing the method, taking measurements and analyzing the results.
The scenario describes the programming of a function performing a binary search on an array of characters. It consists of 19 steps, including 10 steps where the source code is modified and 9 steps where the parameters are changed.

@@

%vRozen: Parking some ideas here about the relationship between stack recordings, cause-and-effect chains and different live programming features. It may be this fits better in related work or future work than the discussion.
The LiveRec approach introduces the minimal data structure, the stack recording, for supporting probes.
This live programming feature is limited to analyzing functions and methods.
Of course, other liveness features exist that require a more elaborate structure and debugging instrumentation.
In particular, omniscient debugging (also known as back in time debugging), supports exploratory programming with what-if scenarios through record, replay, undo and redo functionality.
The modification of running programs requires an explicit representation of change that can augment and integrate with the stack recording for supporting these features.
For instance, the Cascade framework represents change as cause-and-effect chains (based on edit operations) that can update both a live program's syntax and its run-time state.
Future work can further investigate how the generic debugging instrumentation can be extended for supporting additional live programming features.
